---
title: "Package MKpower"
author: "Matthias Kohl"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{MKpower}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{utf8}
---


## Introduction  
The package includes functions for power analysis and sample size calculation. 
Examples are Welch and Hsu t-Tests, Wilcoxon Rank Sum and Signed Rank Tests 
as well as power and sample size required for the evaluation of a diagnostic 
test(-system).

We first load the package.
```{r}
library(MKpower)
```


## Single Proportion


## Welch Two-Sample t-Test
For computing the sample size of the Welch t-test, we only consider the situation 
of equal group size (balanced design).
```{r}
## identical results as power.t.test, since sd = sd1 = sd2 = 1
power.welch.t.test(n = 20, delta = 1)
power.welch.t.test(power = .90, delta = 1)
power.welch.t.test(power = .90, delta = 1, alternative = "one.sided")

## sd1 = 0.5, sd2 = 1
power.welch.t.test(delta = 1, sd1 = 0.5, sd2 = 1, power = 0.9)
```


## Hsu Two-Sample t-Test
For computing the sample size of the Hsu t-test, we only consider the situation 
of equal group size (balanced design).
```{r}
## slightly more conservative than Welch t-test
power.hsu.t.test(n = 20, delta = 1)
power.hsu.t.test(power = .90, delta = 1)
power.hsu.t.test(power = .90, delta = 1, alternative = "one.sided")

## sd1 = 0.5, sd2 = 1
power.welch.t.test(delta = 0.5, sd1 = 0.5, sd2 = 1, power = 0.9)
power.hsu.t.test(delta = 0.5, sd1 = 0.5, sd2 = 1, power = 0.9)
```


## Wilcoxon Rank Sum and Signed Rank Tests
For computing the sample size of these tests, we offer a function that 
performs Monte-Carlo simulations to determine their (empirical) power.
```{r}
rx <- function(n) rnorm(n, mean = 0, sd = 1) 
ry <- function(n) rnorm(n, mean = 0.5, sd = 1) 
## two-sample
sim.ssize.wilcox.test(rx = rx, ry = ry, n.max = 100, iter = 1000)
sim.ssize.wilcox.test(rx = rx, ry = ry, n.min = 65, n.max = 70, 
                      step.size = 1, iter = 1000, BREAK = FALSE)
power.t.test(delta = 0.5, power = 0.8)

## one-sample
sim.ssize.wilcox.test(rx = ry, n.max = 100, iter = 1000, type = "one.sample")
sim.ssize.wilcox.test(rx = ry, type = "one.sample", n.min = 33, 
                      n.max = 38, step.size = 1, iter = 1000, BREAK = FALSE)
power.t.test(delta = 0.5, power = 0.8, type = "one.sample")

## two-sample
rx <- function(n) rgamma(n, scale = 2, shape = 1.5)
ry <- function(n) rgamma(n, scale = 4, shape = 1.5) # equal shape
## two-sample
sim.ssize.wilcox.test(rx = rx, ry = ry, n.max = 100, iter = 1000)
sim.ssize.wilcox.test(rx = rx, ry = ry, n.min = 25, n.max = 30, 
                      step.size = 1, iter = 1000, BREAK = FALSE)
```

For pratical applications we recommend to use a higher number of iterations. 
More detailed examples we refer to the help page of the function.


## Two Negative Binomial Rates
When we consider two negative binomial rates, we can compute sample size or
power applying function power.nb.test.
```{r}
## examples from Table III in Zhu and Lakkis (2014)
power.nb.test(mu0 = 5.0, RR = 2.0, theta = 1/0.5, duration = 1, power = 0.8, approach = 1)
power.nb.test(mu0 = 5.0, RR = 2.0, theta = 1/0.5, duration = 1, power = 0.8, approach = 2)
power.nb.test(mu0 = 5.0, RR = 2.0, theta = 1/0.5, duration = 1, power = 0.8, approach = 3)
```


## Diagnostic Test
Given an expected sensitivity and specificity we can compute sample size, 
power, delta or significance level of diagnostic test.
```{r}
## see n2 on page 1202 of Chu and Cole (2007)
power.diagnostic.test(sens = 0.99, delta = 0.14, power = 0.95) # 40
power.diagnostic.test(sens = 0.99, delta = 0.13, power = 0.95) # 43
power.diagnostic.test(sens = 0.99, delta = 0.12, power = 0.95) # 47
```

The sample size planning for developing binary classifiers in case of high
dimensional data, we can apply function ssize.pcc, which is based on the 
probability of correct classification (PCC).
```{r}
## see Table 2 of Dobbin et al. (2008)
g <- 0.1
fc <- 1.6
ssize.pcc(gamma = g, stdFC = fc, nrFeatures = 22000)
```


## sessionInfo
```{r}
sessionInfo()
```
